---
title: "HCUP Data Analysis - National Inpatient Sample"
output:
  html_document:
    df_print: paged
---

## Load Packages
```{r}
library(reshape)
library(stringr)
library(ggplot2)
library(sqldf)
library(ggrepel)
library(cowplot)
library(trend)
library(ggthemes)
library(grid)
library(extrafont)
  # Make sure to run font_import() command in console if not done previously
library(scales)
```

## Assign Functions
```{r}
# Lagpad function 
lagpad <- function(x, k) {
    if (!is.vector(x)) 
        stop('x must be a vector')
    if (!is.numeric(x)) 
        stop('x must be numeric')
    if (!is.numeric(k))
        stop('k must be numeric')
    if (1 != length(k))
        stop('k must be a single number')
    c(rep(NA, k), x)[1 : length(x)]
}

# Graphics Theme function
econ_theme <-  function() {
  theme(plot.title = element_text(family = 'Century Gothic', hjust=0,size=18,margin=margin(-7,0,10,0)),
    plot.subtitle = element_text(family = 'Century Gothic',size=12, margin=margin(-2,0,3,0)),
    plot.caption = element_text(family = 'Century Gothic', size=9, hjust = 0,margin=margin(34,0,-45,0)),  
    axis.text = element_text(face='bold', family = 'Century Gothic'),
    axis.title.x = element_text(family = 'Century Gothic',face = 'bold', margin=margin(12,0,-12,0)),
    axis.title.y = element_text(family = 'Century Gothic',face = 'bold', margin=margin(0,12,0,0)),
    plot.margin = margin(1,1.5,2,0.5, 'cm'))
}
```


## 1) Analyze National Inpatient Sample - Major Diagnositic Categories (MDC)
Data was sourced from HCUPnet - AHRQ's self-service query tool for simple aggregate analysis of HCUP data assets. This workbook uses the National Inpatient Sample, a clustered sample of inpatient claims throughout the US.
Query options include: 1) Inpatient Setting, 2) All Discharge Data. 3) By by Year. A 2nd query was run with same parameters by MDC. HMO Enrollment data was also obtained.

Data output was formatted to remove total number of discharges and any notes, leaving year and rate of discharges per 100,000. 
```{r}
# Infile data
MDC_filepath <- "X:\\General Folder\\HC_Data\\HCUP_2018\\HCUP_MDC.txt"
MDCse_filepath <- "X:\\General Folder\\HC_Data\\HCUP_2018\\HCUP_MDCse.txt"
MDC_ALOS_filepath <- "X:\\General Folder\\HC_Data\\HCUP_2018\\MDC_ALOS.txt"
MDCse_ALOS_filepath <- "X:\\General Folder\\HC_Data\\HCUP_2018\\MDC_ALOSse.txt"
MDC_all_filepath <- "X:\\General Folder\\HC_Data\\HCUP_2018\\Inpat_discharge_alos_ttl.txt"

MDC <- read.table(MDC_filepath,header= TRUE, sep = '\t', na.strings="*")
MDCse <- read.table(MDCse_filepath,header= TRUE, sep = '\t', na.strings="*")
MDC_ALOS <- read.table(MDC_ALOS_filepath,header= TRUE, sep = '\t', na.strings="*")
MDC_ALOSse <- read.table(MDCse_ALOS_filepath,header= TRUE, sep = '\t', na.strings="*")
MDC_all <- read.table(MDC_all_filepath,header= TRUE, sep = '\t', na.strings="*")

MDC_Datasets <- c("MDC", "MDCse", "MDC_ALOS", "MDC_ALOSse")

# Remove MDC 0 - this is all noise
MDC <- MDC[!(MDC$MDC=="MDC: 0 Principal Dx Can Not Be Assigned To Mdc (Invalid Or Pre Mdc)"),]
MDCse <- MDCse[!(MDCse$MDC=="MDC: 0 Principal Dx Can Not Be Assigned To Mdc (Invalid Or Pre Mdc)"),]
MDC_ALOS <- MDC_ALOS[!(MDC_ALOS$MDC=="MDC: 0 Principal Dx Can Not Be Assigned To Mdc (Invalid Or Pre Mdc)"),]
MDC_ALOSse <- MDC_ALOSse[!(MDC_ALOSse$MDC=="MDC: 0 Principal Dx Can Not Be Assigned To Mdc (Invalid Or Pre Mdc)"),]

# Reshape Data, and format
MDC_long <- melt(MDC, id="MDC")
colnames(MDC_long) <- c("MDC","year","discharge")

MDC_selong <- melt(MDCse, id="MDC")
colnames(MDC_selong) <- c("MDC","year","discharge_se")

MDC_ALOS <- melt(MDC_ALOS, id="MDC")
colnames(MDC_ALOS) <- c("MDC","year","ALOS")

MDC_ALOSse <- melt(MDC_ALOSse, id="MDC")
colnames(MDC_ALOSse) <- c("MDC","year","ALOS_se")

MDC_long$year <- substr(MDC_long$year,2,5)
MDC_long$year <- as.numeric(MDC_long$year)

MDC_selong$year <- substr(MDC_selong$year,2,5)
MDC_selong$year <- as.numeric(MDC_selong$year)

MDC_ALOS$year <- substr(MDC_ALOS$year,2,5)
MDC_ALOS$year <- as.numeric(MDC_ALOS$year)

MDC_ALOSse$year <- substr(MDC_ALOSse$year,2,5)
MDC_ALOSse$year <- as.numeric(MDC_ALOSse$year)

MDC_long$short_name <- substr(MDC_long$MDC,1,7)

# Merge Data into one long data set
MDC_long <- sqldf('SELECT a.*, b.discharge_se
              FROM MDC_long AS A INNER JOIN MDC_selong AS B 
              ON a.MDC = b.MDC
              AND a.year = b.year
             ')

MDC_long <- sqldf('SELECT a.*, b.ALOS
              FROM MDC_long AS A INNER JOIN MDC_ALOS AS B 
              ON a.MDC = b.MDC
              AND a.year = b.year
             ')

MDC_long <- sqldf('SELECT a.*, b.ALOS_se
              FROM MDC_long AS A INNER JOIN MDC_ALOSse AS B 
              ON a.MDC = b.MDC
              AND a.year = b.year
             ')


# Calculate 95% CI bounds from SE
MDC_long$discharge_CI <- 1.96*MDC_long$discharge_se
MDC_long$ALOS_CI <- 1.96*MDC_long$ALOS_se

# Calculate percent change label for discharge total table
MDC_all$discharge_rate_perc_change <- paste(as.character(signif((MDC_all$Discharge_Rate - (lagpad(MDC_all$Discharge_Rate, 1))) / (lagpad(MDC_all$Discharge_Rate, 1))*100, digits=2)),"%")

MDC_all$discharge_rate_perc_change[MDC_all$discharge_rate_perc_change == "NA %"] <- ""

MDC_all$discharge_perc_change <- paste(as.character(signif((MDC_all$Total_Discharges - (lagpad(MDC_all$Total_Discharges, 1))) / (lagpad(MDC_all$Total_Discharges, 1))*100, digits=2)),"%")

MDC_all$discharge_perc_change[MDC_all$discharge_perc_change == "NA %"] <- ""

# Calculate Bed Days
MDC_all$bed_days <- MDC_all$Total_Discharges * MDC_all$LOS

```


Plot Total Inpatient Discharge Rate per 100,000
```{r fig.width=10, fig.height=15}

# Plot Discharge Rate
plot <- ggplot(MDC_all, aes(x=Year, y=Discharge_Rate, label = discharge_rate_perc_change
                     )) + 
  geom_line(col='#014d64', size = 1.2) + 
  geom_point(col='#014d64', size = 2) + 
  geom_text(check_overlap = TRUE, angle = 65, nudge_y = 125, nudge_x = 0, family = 'Century Gothic') + 
  annotate("text", label = "Labels = % Year-over-Year Change", x = 1997, y = 11650, 
           size = 5, colour = "black", family = 'Century Gothic') +  
  labs(title = 'U.S. Inpatient Discharges per 100,000', 
       subtitle = 'HMO enrollment or other changes in healthcare financing could help explain changes in inpatient utilization',
       caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
  labs(x='Year', y='Inpatient Discharges per 100,000 Population') + 
  scale_y_continuous(label=comma) + 
  theme_economist_white() + econ_theme()

print(plot + theme(legend.position = "none"))
grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph


```

Plot All Inpatient Discharges
```{r fig.width=10}

# Plot all Discharges
plot <- ggplot(MDC_all, aes(x=Year, y=Total_Discharges, label = discharge_perc_change
                     )) + 
  geom_line(col='#014d64', size = 1.2) + 
  geom_point(col='#014d64', size = 2) + 
  geom_text(check_overlap = TRUE, angle = 65, nudge_y = 700000, nudge_x = 0, family = 'Century Gothic') + 
  annotate("text", label = "Labels = % Year-over-Year Change", x = 1998, y = 32500000, 
           size = 5, colour = "black", family = 'Century Gothic') +  
  labs(title = 'Total U.S. Inpatient Discharges', 
       subtitle = 'The total number of U.S. acute inpatient discharges have been in decline since 2010', 
       caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
  labs(x='Year', y='Acute Inpatient Discharges') + 
  scale_y_continuous(label=comma) + 
  theme_economist_white() + econ_theme()

print(plot + theme(legend.position = "none"))
grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph

```

Plot Total ALOS
```{r fig.width=10}

# Plot Discharge Rate
plot <- ggplot(MDC_all, aes(x=Year, y=LOS
                     )) + 
  geom_line(col='#014d64', size = 1.2) + 
  geom_point(col='#014d64', size = 2) + 
  labs(title = 'Average Length of Stay', 
       subtitle = 'Length of hospitals stays decreased sharply prior to 2000', 
       caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
  labs(x='Year', y='Average Length of Stay (Days)') + 
  theme_economist_white() + econ_theme()

print(plot + theme(legend.position = "none"))
grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph

```

Plot Bed Days
```{r fig.width=10}

# Plot Discharge Rate
plot <- ggplot(MDC_all, aes(x=Year, y=bed_days
                     )) + 
  geom_line(col='#014d64', size = 1.2) + 
  geom_point(col='#014d64', size = 2) + 
  labs(title = 'U.S. Bed Days (Estimated)', 
       subtitle = 'Averge Length of Stay * Discharges - likely overestimate of actual bed days', 
       caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
  labs(x='Year', y='Bed Days') + 
  scale_y_continuous(label=comma) + 
  theme_economist_white() + econ_theme()

print(plot + theme(legend.position = "none"))
grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph

```

Plot Discharge Rate by MDC
```{r fig.width=10, fig.height=22}
# Set Colors
cbbPalette <- c("#000000", "#666600", "#000033", "#00FF33", "#CC6600", "#FF6699", "#D55E00", "#CC79A7", "#99CC00", "#00CC99", "#00FF33", "#00CCFF", "#0033FF", "#3300CC", "#9900FF", "#CC33CC", "#FFCC33", "#003333", "#666600", "#FF9933", "#3399FF", "#CC0099", "#CC0000", "#330033", "#0000FF", "#FFFF99")

# Summary Plot, to get sense of scale
MDC_long$end_label <- MDC_long$short_name
MDC_long$end_label[MDC_long$year < 2015] <- ""

plot <- ggplot(MDC_long, aes(x=year, y=discharge, group=MDC, color=MDC, label = end_label
                     )) + 
  geom_line(position = 'jitter', size = 1) + 
  geom_text_repel(segment.color = "black", family = 'Century Gothic') + 
  scale_colour_manual(values=cbbPalette) + 
  labs(title = 'Inpatient Discharges per 100k, by Major Diagnostic Category (MDCs)', 
       subtitle = 'With some exception, the rate of discharges for most Major Diagnostic Categories have been in decline', 
       caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
  labs(x='Year', y='Inpatient Discharges per 100,000 Population') + 
  scale_y_continuous(label=comma) + 
  theme_economist_white() + econ_theme()

print(plot + theme(legend.position = "bottom") + guides(col = guide_legend(nrow=25)))
grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph

```

Plot ALOS by MDC
```{r fig.width=10, fig.height=12}
plot <- ggplot(MDC_long, aes(x=year, y=ALOS, group=MDC, color=MDC, label = end_label
                     )) + 
  geom_line(position = 'jitter', size = 1) + 
  geom_text_repel(segment.color = "black", family = 'Century Gothic') + 
  scale_colour_manual(values=cbbPalette) + 
  labs(title = 'Average Length of Stay (ALOS), by Major Diagnostic Category (MDCs)', 
       subtitle = 'With some exception, average length of stay has decreased over time', 
       caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
  labs(x='Year', y='Average length of stay (ALOS)') + 
  scale_y_continuous(label=comma) + 
  theme_economist_white() + econ_theme()

print(plot + theme(legend.position = "none") + guides(col = guide_legend(nrow=25)))
grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph
```

Plot Discharge Rate and ALOS by MDC w. 95% Confidence Intervals
```{r, fig.width=10, fig.height=12}
# Plot by MDC, with 95% CIs
MDCs <- unique(MDC_long$MDC)

# Get end and start of ALOS for label
MDC_long$ALOS_label <- signif(MDC_long$ALOS,2)
MDC_long$ALOS_label[MDC_long$year < 2015 & MDC_long$year > 1993] <- ""


for (i in seq_along(MDCs)) {
  
  plot <- ggplot(subset(MDC_long, MDC_long$MDC==MDCs[i]), aes(x=year, y=discharge, group=MDC
                       )) + 
    geom_line(col='#014d64', linetype = 2, size = 1) + 
    geom_ribbon(alpha=0.2, linetype = 2, aes(ymin=discharge-discharge_CI, ymax=discharge+discharge_CI)) + 
    ggtitle(paste(MDCs[i]," w. 95% CI")) + 
    scale_y_continuous(label=comma) +
    labs(x='Year', y='Inpatient discharges per 100,000 population') + 
    theme_economist_white() + econ_theme()

  
  plot2 <- ggplot(subset(MDC_long, MDC_long$MDC==MDCs[i]), aes(x=year, y=ALOS, group=MDC, label=ALOS_label
                       )) + 
    geom_line(col='#014d64', linetype = 2, size = 1) + 
    geom_ribbon(alpha=0.2, aes(ymin=ALOS-ALOS_CI, ymax=ALOS+ALOS_CI), linetype = 2) + 
    geom_text_repel(segment.color = "black", nudge_y=1, family = 'Century Gothic') + 
    ggtitle("Average Length of Stay (ALOS) w. 95% CI") + 
    scale_y_continuous(limits = c(0, 16)) +
    labs(x='Year', y='Average length of stay (ALOS)',
    caption = 'Source: Agency for Healthcare Quality - HCUP National Inpatient Sample.') + 
    theme_economist_white() + econ_theme()

  # Cowplot grids together
  print(cowplot::plot_grid(plot, plot2, nrow=2, align="v", axis='1', rel_heights = c(.6,.4)))
  grid.rect(x=0.026, y=0.9,hjust = 1,vjust=0,gp=gpar(fill='#014d64',lwd=0))# Add Blue rectangle to the graph

}

```

Summary
```{r}

# Pull Summary Stats
MDC$Diff <- MDC$X2015 - MDC$X1993
MDC$perc_change <- MDC$Diff/MDC$X2015
MDC <- sqldf('SELECT a.*, b.X1993 AS X1993se, b.X2015 AS X2015se
              FROM MDC AS A LEFT JOIN MDCse AS B
              ON a.MDC = b.MDC
          ')

# Mann-Kendall Test, since 2010

MDC10 <- subset(MDC_long,year>2009)

for (i in seq_along(MDCs)) {
  x <- subset(MDC10, MDC10$MDC==MDCs[i])
  print(MDCs[i])
  x <- as.ts(x[c("discharge")])
  pval <- mk.test(x, alternative = "two.sided")
  print(pval$p.value)
}

```

